\documentclass[11pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,fit,backgrounds}

% Code listing style
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!10},
    keywordstyle=\color{blue},
    commentstyle=\color{green!60!black},
    stringstyle=\color{red!60!black},
    numbers=left,
    numberstyle=\tiny\color{gray},
    numbersep=5pt,
    showstringspaces=false
}

% Custom commands
\newcommand{\cognitex}{\textsc{Cognitex}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}

\title{\cognitex{}: A Multi-Tier Personal Agent Architecture for Cognitive Load Management Through Semantic Knowledge Graphs, Autonomous Graph Orchestration, State-Aware Executive Function, and Behavioral Learning}

\author{
    Chris Sainsbury\\
    Independent Researcher\\
    United Kingdom\\
    \texttt{chris@sainsbury.im}
}

\date{January 2025}

\begin{document}

\maketitle

\begin{abstract}
We present \cognitex{}, a personal agent architecture designed to reduce cognitive overhead for knowledge workers by unifying email, calendar, document, and code repository data into a hybrid semantic knowledge graph. The system implements a hierarchical large language model (LLM) architecture with distinct reasoning and execution tiers, combined with a three-level risk stratification framework (read-only, auto-execute, approval-required) that balances autonomous operation with human agency.

A key innovation is the \textbf{Digital Twin} capability: an autonomous agent that learns the user's communication style from sent emails and can draft replies, compile context packs for upcoming meetings, and suggest focus time blocks for stalled projects. The agent proactively monitors graph health, creates connections between related entities, and surfaces items requiring human attention via Discord notifications. A dedicated review dashboard enables users to approve, edit, or discard agent-generated drafts before execution.

The architecture introduces a sophisticated \textbf{Executive Function Layer} that models user cognitive state through six operating modes (Deep Focus, Fragmented, Overloaded, Avoidant, Hyperfocus, Transition) with continuous signals for fatigue, interruption pressure, and focus. A \textbf{Temporal Energy Model} captures diurnal patterns---peak hours, post-lunch dips, evening decline---enabling state-aware task suggestions. Clinical sessions (high-drain events) trigger automatic recovery mode for the remainder of the day, with the system learning from task completion observations to refine its understanding of user capacity patterns.

A \textbf{Behavioral Learning System} extracts preference rules from user feedback, tracks proposal acceptance patterns, calibrates duration estimates, predicts deferral risk, and injects learned guidelines back into agent prompts. Rules follow a lifecycle from candidate through validated to deprecated, with automatic retirement of underperforming patterns.

The architecture integrates three complementary knowledge representations: a Neo4j property graph capturing entities and relationships, PostgreSQL with pgvector extensions for 768-dimensional semantic embeddings, and Redis-backed working memory for session context. The agent employs a ReAct-style (Reason + Act) planning loop with decision trace capture, and externalized prompt templates for iterative refinement. \cognitex{} represents a reference implementation for trustworthy personal AI systems that maintain transparency while providing meaningful cognitive assistance through both reactive and proactive behaviors.
\end{abstract}

\noindent\textbf{Keywords:} Personal agents, Knowledge graphs, Large language models, Cognitive load management, Semantic search, ReAct agents, Autonomous orchestration, Human-AI collaboration, Behavioral learning, Executive function modeling

\section{Introduction}

\subsection{Motivation}

Knowledge workers face an increasingly fragmented digital environment characterized by multiple communication channels, calendaring systems, document repositories, and project management tools. This fragmentation imposes substantial cognitive overhead: the mental effort required to track commitments, maintain relationship context, and synthesize information across disparate sources. Studies estimate that knowledge workers spend 28\% of their workweek managing email alone \citep{mckinsey2012social}, with significant additional time lost to context switching between applications.

While general-purpose AI assistants have demonstrated impressive capabilities in natural language understanding and generation, they typically operate without persistent context about the user's specific professional network, commitments, and communication patterns. This limitation necessitates repeated context provision, undermining the efficiency gains such systems might otherwise provide. Furthermore, most AI assistants are purely \emph{reactive}---they respond to user queries but do not proactively maintain or organize information.

Critically, existing systems lack awareness of user cognitive state. They suggest demanding tasks when the user is exhausted, interrupt deep work with routine notifications, and fail to learn which suggestions are accepted versus rejected. A truly effective personal agent must model user capacity and adapt its behavior accordingly.

\subsection{Contributions}

This paper presents \cognitex{}, a personal agent system that addresses these challenges through several novel contributions:

\begin{enumerate}
    \item \textbf{Hybrid Knowledge Representation}: We integrate property graphs (Neo4j), vector embeddings (PostgreSQL/pgvector), and key-value caching (Redis) to capture complementary aspects of personal knowledge---relationships, semantic similarity, and session context respectively.

    \item \textbf{Digital Twin Agent}: An autonomous agent that learns the user's writing style from sent emails and acts on their behalf---drafting email replies, compiling context packs for meetings, and suggesting focus time blocks---with a dedicated review dashboard for approval before execution.

    \item \textbf{Autonomous Graph Orchestration}: An always-running agent loop that proactively monitors graph state, identifies connection opportunities (documents matching project names, repositories matching projects, orphaned tasks), creates links automatically, and surfaces ambiguous items for human review via Discord notifications.

    \item \textbf{Risk-Stratified Tool Execution}: Our framework categorizes agent capabilities into three risk tiers (\texttt{READONLY}, \texttt{AUTO}, \texttt{APPROVAL}), enabling autonomous operation for safe actions while preserving human oversight for consequential decisions.

    \item \textbf{Executive Function Layer}: A comprehensive cognitive state model with six operating modes, continuous signals (fatigue, interruption pressure, focus score), mode-specific rules governing task eligibility and notification filtering, and an interruption firewall that queues non-urgent items during focused work.

    \item \textbf{Temporal Energy Model}: A diurnal energy curve modeling peak hours (typically 8--11am), post-lunch dips, and evening decline, with support for learning personalized patterns from task completion observations.

    \item \textbf{Clinical Session Detection}: Automatic identification of high-drain calendar events (marked with \texttt{[CLINICAL]} tags) that trigger rest-of-day recovery mode, preventing the agent from suggesting demanding tasks after cognitively exhausting activities.

    \item \textbf{Behavioral Learning System}: Extraction of preference rules from user feedback with lifecycle management (candidate $\rightarrow$ validated $\rightarrow$ deprecated), proposal pattern learning, duration calibration, deferral risk prediction, and injection of learned guidelines into agent prompts.

    \item \textbf{State-Aware Task Suggestions}: The autonomous agent respects current cognitive state, filtering tasks by mode-appropriate friction levels and suggesting rescheduling of high-demand tasks to peak energy hours.

    \item \textbf{Decision Trace Learning}: We capture complete decision contexts including triggers, proposed actions, and user feedback, enabling retrieval-augmented generation from similar historical decisions and providing infrastructure for behavioral fine-tuning.

    \item \textbf{Externalized Prompt Architecture}: System prompts are stored as markdown files with templated variables, enabling rapid iteration on agent behavior without code changes.

    \item \textbf{Multi-Modal Triggering}: The agent responds to scheduled events (morning briefings, periodic monitoring), external triggers (new emails, calendar changes, Drive file updates), autonomous cycles (15-minute graph health checks), and direct user interaction through multiple interfaces (CLI, Discord, web dashboard, REST API).

    \item \textbf{Coding Session Ingestion}: We introduce a distributed session sync architecture that captures AI coding assistant interactions (Claude Code, etc.) from multiple machines, extracts summaries and decisions via LLM, and links sessions to projects for rich development context. A standalone \texttt{cognitex-sync} package enables remote machines to push session data to the central server.

    \item \textbf{Human-in-the-Loop Entity Linking}: A fuzzy matching interface for linking unstructured data to structured entities, allowing users to review and approve suggested matches before committing to the knowledge graph.
\end{enumerate}

\subsection{Paper Organization}

\Cref{sec:related} reviews related work in personal agents, knowledge graphs, and LLM-based systems. \Cref{sec:architecture} presents the system architecture, including data models and service integrations. \Cref{sec:agent} details the agent subsystem, covering the planner, executors, tools, and memory systems. \Cref{sec:autonomous} describes the autonomous graph orchestration layer. \Cref{sec:executive} covers the executive function layer including state modeling, temporal energy patterns, and interruption management. \Cref{sec:learning} presents the behavioral learning system. \Cref{sec:graph} describes the semantic graph construction and query patterns. \Cref{sec:chunking} covers the document chunking and embedding pipeline. \Cref{sec:triggers} discusses the trigger and notification infrastructure. \Cref{sec:web} presents the web dashboard interface. \Cref{sec:evaluation} presents preliminary evaluation results. \Cref{sec:discussion} discusses limitations and future directions, and \Cref{sec:conclusion} concludes.

\section{Related Work}
\label{sec:related}

\subsection{Personal Information Management}

Personal information management (PIM) systems have evolved from simple contact databases to sophisticated knowledge organization tools. Semantic desktop initiatives \citep{sauermann2005semantic} proposed treating personal data as a knowledge graph, though adoption was limited by the manual effort required for ontology construction. More recent approaches leverage machine learning for automatic relationship extraction \citep{dong2014knowledge}, though typically focusing on enterprise rather than personal contexts.

\subsection{LLM-Based Agents}

The emergence of capable large language models has catalyzed research into autonomous agents. ReAct \citep{yao2022react} demonstrated the effectiveness of interleaved reasoning and action traces for tool-using agents. Subsequent work has explored tool learning \citep{qin2023toolllm}, self-reflection \citep{shinn2023reflexion}, and multi-agent collaboration \citep{park2023generative}. However, most systems target general-purpose assistance rather than the specialized domain of personal productivity with its requirements for persistent context and relationship awareness.

\subsection{Proactive Agent Behavior}

Traditional agent systems are predominantly reactive, responding only to explicit user requests. Recent work has explored proactive behaviors including anticipatory assistance \citep{yorke2012proactive} and context-aware notifications. Our autonomous orchestration layer extends this paradigm by continuously monitoring a knowledge graph and taking corrective actions to maintain graph health and coherence.

\subsection{Cognitive State Modeling}

Research in human-computer interaction has long recognized the importance of user state in system design. Interruption management systems \citep{horvitz1999principles} model attention and interruptibility. More recent work explores physiological signals for state inference. Our executive function layer builds on this tradition, providing a structured model of cognitive capacity that governs agent behavior.

\subsection{Knowledge Graph Construction}

Automatic knowledge graph construction from unstructured text has received extensive attention \citep{nickel2016review}. Entity recognition, relation extraction, and entity linking form the core pipeline. For personal knowledge graphs specifically, challenges include handling the diversity of personal data sources and maintaining privacy constraints. Our approach differs by leveraging LLM capabilities for flexible entity extraction while maintaining structured schemas for core entity types.

\subsection{Semantic Search and Retrieval}

Dense retrieval methods using learned embeddings have largely superseded sparse retrieval for semantic search tasks \citep{karpukhin2020dense}. The integration of vector databases with traditional relational systems enables hybrid queries combining structured filters with semantic similarity. Our architecture employs this pattern, using PostgreSQL's pgvector extension for embedding storage alongside structured metadata.

\section{System Architecture}
\label{sec:architecture}

\subsection{Overview}

\cognitex{} implements a multi-tier architecture comprising data services, integration services, agent components, autonomous orchestration, executive function management, behavioral learning, and user interfaces. The system operates as a set of containerized microservices orchestrated via Docker Compose. \Cref{fig:architecture} illustrates the high-level system design.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=0.7cm,
    box/.style={rectangle, draw, rounded corners, minimum width=1.8cm, minimum height=0.7cm, align=center, font=\small},
    bigbox/.style={rectangle, draw, rounded corners, minimum width=4cm, minimum height=1cm, align=center},
    layer/.style={rectangle, draw, dashed, rounded corners, inner sep=0.25cm},
    arrow/.style={->, >=stealth, thick}
]

% User Interfaces Layer
\node[box] (cli) {CLI\\(Typer)};
\node[box, right=0.4cm of cli] (discord) {Discord\\Bot};
\node[box, right=0.4cm of discord] (web) {Web\\Dashboard};
\node[box, right=0.4cm of web] (api) {REST API\\(FastAPI)};
\node[layer, fit=(cli)(discord)(web)(api), label=above:User Interfaces] (ui_layer) {};

% Agent System Layer
\node[box, below=1.3cm of cli] (planner) {Planner};
\node[box, right=0.2cm of planner] (executors) {Executors};
\node[box, right=0.2cm of executors] (tools) {Tools\\(27)};
\node[box, right=0.2cm of tools] (autonomous) {Autonomous\\Agent};
\node[box, below=0.4cm of planner, xshift=0.3cm] (memory) {Memory\\Systems};
\node[box, right=0.2cm of memory] (executive) {Executive\\Layer};
\node[box, right=0.2cm of executive] (learning) {Learning\\System};
\node[box, below=0.4cm of memory, xshift=0.8cm] (triggers) {Trigger System};
\node[layer, fit=(planner)(executors)(tools)(autonomous)(memory)(executive)(learning)(triggers), label=above:Agent System] (agent_layer) {};

% Integration Services Layer
\node[box, below=1.3cm of triggers, xshift=-2.5cm] (gmail) {Gmail};
\node[box, right=0.2cm of gmail] (calendar) {Calendar};
\node[box, right=0.2cm of calendar] (drive) {Drive};
\node[box, right=0.2cm of drive] (github) {GitHub};
\node[box, right=0.2cm of github] (llm) {LLM};
\node[layer, fit=(gmail)(calendar)(drive)(github)(llm), label=above:Integration Services] (int_layer) {};

% Data Services Layer
\node[box, below=1.3cm of gmail, xshift=0.5cm] (postgres) {PostgreSQL\\+pgvector};
\node[box, right=0.4cm of postgres] (neo4j) {Neo4j\\Graph};
\node[box, right=0.4cm of neo4j] (redis) {Redis\\Cache};
\node[layer, fit=(postgres)(neo4j)(redis), label=above:Data Services] (data_layer) {};

% Arrows
\draw[arrow] (ui_layer) -- (agent_layer);
\draw[arrow] (agent_layer) -- (int_layer);
\draw[arrow] (int_layer) -- (data_layer);

\end{tikzpicture}
\caption{High-level system architecture of \cognitex{} showing the four primary layers: user interfaces (CLI, Discord, Web Dashboard, REST API), agent system (including autonomous orchestration, executive function layer, and learning system), integration services, and data services.}
\label{fig:architecture}
\end{figure}

\subsection{Data Services}

\subsubsection{PostgreSQL with pgvector}

PostgreSQL serves as the primary relational store, extended with the pgvector extension for efficient similarity search over dense embeddings. The schema comprises 25+ tables organized into functional groups:

\paragraph{Task Management:}
\begin{itemize}
    \item \texttt{tasks}: Core task storage with status (pending, in\_progress, completed), energy cost estimates (1--5 scale), due dates, priority levels, estimated/actual duration, deferral counts, and foreign keys to source entities
    \item \texttt{goals}: Hierarchical goal management supporting quarterly and yearly timeframes with OKR-style key results stored as JSONB
    \item \texttt{projects}: Project containers linking goals to executable tasks with status tracking (active, completed, on\_hold, cancelled)
\end{itemize}

\paragraph{Semantic Search:}
\begin{itemize}
    \item \texttt{embeddings}: Universal embedding storage with entity\_type discriminator, 768-dimensional vectors, and content hashes for deduplication
    \item \texttt{document\_chunks}: Chunked document segments with positional metadata
    \item \texttt{code\_content}: GitHub source code with repository and path metadata
\end{itemize}

\paragraph{Agent Operations:}
\begin{itemize}
    \item \texttt{agent\_actions}: Complete log of autonomous agent actions including action\_type, source (agent, trigger, user), status, summary, and detailed JSONB metadata
    \item \texttt{decision\_traces}: Complete decision context including trigger type, proposed action, final action, user feedback, and quality scores
    \item \texttt{task\_proposals}: Agent-proposed tasks with status (pending, approved, rejected, expired) and user feedback
    \item \texttt{communication\_patterns}: Learned per-contact preferences
\end{itemize}

\paragraph{Learning System:}
\begin{itemize}
    \item \texttt{preference\_rules}: Extracted behavioral patterns with conditions, confidence scores, lifecycle status (candidate, validated, deprecated), application counts, and success rates
    \item \texttt{learned\_patterns}: Cached patterns by type (proposal\_acceptance, temporal, deferral, duration) with confidence and sample size
    \item \texttt{task\_timing}: Duration estimates vs. actuals for calibration
    \item \texttt{deferral\_analysis}: Task deferral history with reasons
\end{itemize}

\paragraph{State Observations:}
\begin{itemize}
    \item \texttt{state\_observations}: Task outcomes with cognitive state context (mode, fatigue, hour, post-clinical flag) for temporal pattern learning
    \item \texttt{interruption\_events}: Logged interruptions with mode context for firewall tuning
\end{itemize}

The embedding index employs HNSW (Hierarchical Navigable Small World) for approximate nearest neighbor search:

\begin{lstlisting}[language=SQL]
CREATE INDEX idx_embeddings_vector
ON embeddings USING hnsw (embedding vector_cosine_ops)
WITH (m = 16, ef_construction = 64);
\end{lstlisting}

This configuration provides $O(\log n)$ query complexity with high recall at the cost of $O(n \log n)$ index construction time.

\subsubsection{Neo4j Property Graph}

Neo4j stores the semantic knowledge graph with 15+ primary node types and 20+ relationship types. \Cref{tab:node_types} summarizes the core entity nodes.

\begin{table}[t]
\centering
\caption{Core node types in the \cognitex{} knowledge graph.}
\label{tab:node_types}
\begin{tabular}{lll}
\toprule
\textbf{Node Type} & \textbf{Key Properties} & \textbf{Description} \\
\midrule
Person & email, name, org, role & Contacts and collaborators \\
Email & gmail\_id, subject, date & Email messages \\
Event & gcal\_id, title, start, drain\_level & Calendar entries \\
Task & id, title, status, priority, due & Action items \\
Document & drive\_id, name, mime\_type & Drive files \\
Repository & id, name, full\_name & GitHub repositories \\
Project & id, title, status, description & Project containers \\
Goal & id, title, timeframe, target\_date & Strategic objectives \\
Chunk & id, chunk\_index, summary & Document segments \\
Topic & name & Extracted themes \\
Concept & name & Domain concepts \\
EmailDraft & id, to, subject, body, status & Agent-drafted replies \\
ContextPack & id, title, summary, key\_points & Meeting briefings \\
SuggestedBlock & id, title, duration, status & Focus time suggestions \\
StateSnapshot & id, mode, fatigue\_level, created\_at & Cognitive state captures \\
CodingSession & session\_id, project\_path, summary & AI coding assistant sessions \\
\bottomrule
\end{tabular}
\end{table}

The graph captures relationships including:
\begin{itemize}
    \item \textbf{Communication}: \texttt{SENT\_BY}, \texttt{RECEIVED\_BY}, \texttt{MENTIONS}
    \item \textbf{Ownership}: \texttt{OWNS}, \texttt{SHARED\_WITH}, \texttt{CREATED\_BY}
    \item \textbf{Hierarchy}: \texttt{BELONGS\_TO}, \texttt{PART\_OF}, \texttt{DERIVED\_FROM}
    \item \textbf{Semantic}: \texttt{COVERS} (topic), \texttt{ABOUT} (concept), \texttt{REFERENCES}
    \item \textbf{Inferred}: \texttt{WORKS\_WITH}, \texttt{ASSIGNED\_TO}
    \item \textbf{State}: \texttt{CAPTURED\_STATE}, \texttt{STATE\_DURING}
    \item \textbf{Development}: \texttt{DEVELOPS} (session to project), \texttt{WORKS\_ON} (session to repository)
\end{itemize}

\subsubsection{Redis}

Redis provides five functions: (1) working memory with 24-hour TTL for session context, (2) approval queue for staged high-risk actions, (3) pub/sub messaging for event-driven triggers and notifications, (4) deduplication tracking for event processing, and (5) clinical recovery state persistence across service restarts.

\subsection{Integration Services}

\subsubsection{Gmail Service}

The Gmail integration implements OAuth2 authentication with real-time push notifications via Google Cloud Pub/Sub. A pull-based subscription (\texttt{\_gmail\_pubsub\_listener}) receives notifications when new emails arrive, triggering incremental synchronization using Gmail's history API with stored \texttt{history\_id}. Initial synchronization fetches historical messages within a configurable date range. The service supports automatic task completion detection from sent emails.

\subsubsection{Calendar Service}

Google Calendar integration supports full CRUD operations. Events are enriched with computed properties including \texttt{drain\_level} classification:

\begin{itemize}
    \item \textbf{CLINICAL}: Highest drain, identified by \texttt{[CLINICAL]} tag in title; triggers rest-of-day recovery mode
    \item \textbf{HIGH}: Board meetings, presentations, interviews; 1-hour recovery buffer
    \item \textbf{MEDIUM}: Regular meetings
    \item \textbf{LOW}: 1:1s, casual syncs
    \item \textbf{RECOVERY}: Explicitly blocked recovery time
\end{itemize}

\subsubsection{Drive Service}

Google Drive integration indexes document metadata and content with automatic change detection via polling. Content extraction supports Google Docs, PDFs (via pypdf), Microsoft Office documents, and plain text formats. Priority folders can be configured for preferential indexing. The system detects file changes and automatically re-indexes affected documents.

\subsubsection{GitHub Service}

GitHub integration enables code repository indexing with selective targeting of meaningful files (source code, documentation, configuration) while excluding generated artifacts (node\_modules, build outputs, binaries). Repository nodes in the graph can be linked to projects.

\subsubsection{Coding Session Service}

The coding session service captures and indexes interactions from AI coding assistants (Claude Code, Codex, etc.), providing rich development context for projects. The architecture supports both local and remote session ingestion:

\paragraph{Local Session Discovery:} The system scans standard CLI session directories (e.g., \texttt{\textasciitilde/.claude/projects/}) for JSONL session files. Each session is parsed to extract metadata including project path, git branch, timestamps, and message content.

\paragraph{Remote Session Sync:} A standalone \texttt{cognitex-sync} package enables machines without direct database access to push session data to the central server:

\begin{lstlisting}[language=bash]
# Install on remote machine
pip install cognitex-sync

# Configure with server credentials
cognitex-sync configure --server https://cognitex.example.com \
    --api-key <sync_api_key>

# Push sessions (batched to avoid payload limits)
cognitex-sync push
\end{lstlisting}

The sync client discovers sessions, batches them (5 per request to avoid HTTP 413 errors), and pushes to the \texttt{/api/sync/sessions} endpoint with Bearer token authentication.

\paragraph{LLM-Based Summarization:} For sessions containing raw message history, the ingestion pipeline uses an LLM to extract structured summaries:

\begin{lstlisting}[language=Python]
{
    "summary": "Implemented user authentication with JWT",
    "decisions": ["Use bcrypt for password hashing",
                  "Store refresh tokens in Redis"],
    "files_changed": ["src/auth.py", "src/middleware.py"],
    "next_steps": ["Add rate limiting", "Implement logout"],
    "topics": ["authentication", "security", "JWT"],
    "completion_state": "in_progress"
}
\end{lstlisting}

\paragraph{Project Linking:} Sessions are automatically linked to Project and Repository nodes via path matching. A fuzzy matching interface (\texttt{/sessions/link}) enables human-in-the-loop review of suggested matches, calculating similarity scores between session paths and project titles using normalized Levenshtein distance.

\subsubsection{LLM Service}

The LLM service provides a unified interface to inference APIs. Model selection is task-specific:
\begin{itemize}
    \item \textbf{Planning/Reasoning}: DeepSeek V3 (671B parameters, MoE) for complex reasoning
    \item \textbf{Execution}: DeepSeek V3 for structured task completion
    \item \textbf{Embedding}: BAAI/bge-base-en-v1.5 for 768-dimensional semantic vectors
    \item \textbf{Autonomous Agent}: DeepSeek V3 for graph orchestration decisions
\end{itemize}

\section{Agent Subsystem}
\label{sec:agent}

\subsection{Architecture Overview}

The agent subsystem implements a ReAct-style \citep{yao2022react} architecture with hierarchical control flow. The design separates high-level reasoning (what to do and why) from low-level execution (how to accomplish specific tasks), enabling specialized optimization of each tier.

\subsection{Core Agent}

The core agent orchestrates the reasoning loop, managing interactions between the planner, executors, tools, and memory systems. The primary interaction loop proceeds as follows:

\begin{enumerate}
    \item Record user message to working memory
    \item Construct system prompt with available tools
    \item Retrieve relevant historical decisions via RAG
    \item \textbf{Inject learned guidelines from preference rules}
    \item Execute ReAct iterations (maximum 8):
    \begin{itemize}
        \item Generate LLM response with thought/action/observation structure
        \item If action specified, execute tool and capture observation
        \item If response specified, return to user
    \end{itemize}
    \item Record final response to episodic memory
    \item Return response and any approval IDs created
\end{enumerate}

\subsection{Planner}

The planner generates structured action plans from triggers. A plan consists of:

\begin{lstlisting}[language=Python]
@dataclass
class Plan:
    reasoning: str           # Analysis of situation
    steps: list[PlanStep]    # Ordered actions
    user_notification: str   # Optional message
    follow_up: str          # Scheduled check-in
    confidence: float       # Quality score (0-1)
\end{lstlisting}

The planner operates in distinct modes with mode-specific tool access, as shown in \Cref{tab:modes}.

\begin{table}[t]
\centering
\caption{Planner modes and their tool access levels.}
\label{tab:modes}
\begin{tabular}{lll}
\toprule
\textbf{Mode} & \textbf{Description} & \textbf{Tool Access} \\
\midrule
BRIEFING & Morning summary & READONLY, AUTO \\
REVIEW & Evening reflection & READONLY, AUTO \\
MONITOR & Hourly check & READONLY, AUTO \\
PROCESS\_EMAIL & New email analysis & READONLY, AUTO, APPROVAL \\
CONVERSATION & User interaction & READONLY, AUTO, APPROVAL \\
AUTONOMOUS & Graph orchestration & READONLY, AUTO \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Tool Registry}

The system provides 27+ tools organized by risk level:

\paragraph{READONLY Tools (10):} No side effects, always allowed. Includes \texttt{graph\_query}, \texttt{search\_documents}, \texttt{search\_code}, \texttt{get\_calendar}, \texttt{get\_tasks}, \texttt{get\_contact}, \texttt{recall\_memory}, \texttt{get\_projects}, \texttt{get\_goals}, and \texttt{get\_repositories}.

\paragraph{AUTO Tools (14):} Low-risk side effects, executed automatically. Includes \texttt{create\_task}, \texttt{update\_task}, \texttt{create\_project}, \texttt{update\_project}, \texttt{create\_goal}, \texttt{link\_task\_to\_project}, \texttt{link\_project\_to\_goal}, \texttt{link\_document\_to\_project}, \texttt{link\_repository\_to\_project}, \texttt{send\_notification}, \texttt{add\_memory}, \texttt{flag\_for\_review}, and graph mutation tools.

\paragraph{APPROVAL Tools (3):} High-risk, require user confirmation. Includes \texttt{draft\_email}, \texttt{send\_email}, and \texttt{create\_event}.

The risk stratification enables safe autonomous operation while preserving user agency for consequential actions.

\subsection{Memory Systems}

\subsubsection{Working Memory}

Redis-backed short-term storage with 24-hour TTL maintains session context including recent interactions (last 50 messages), current focus topics, and pending approvals.

\subsubsection{Episodic Memory}

PostgreSQL-backed long-term storage captures interactions, decisions, and observations with associated importance scores and entity references.

\subsubsection{Decision Memory}

Specialized storage for learning from feedback captures:
\begin{itemize}
    \item \textbf{Trigger}: What prompted the decision
    \item \textbf{Context}: Full situation snapshot (JSONB)
    \item \textbf{Proposed Action}: What the agent intended
    \item \textbf{Final Action}: What was executed (after edits)
    \item \textbf{Quality Score}: Computed from approval status
\end{itemize}

Quality scoring follows: auto-executed (0.7), approved unchanged (0.9), approved with edits (0.85), rejected (0.3).

\section{Autonomous Graph Orchestration}
\label{sec:autonomous}

\subsection{Motivation}

Traditional agent systems are purely reactive---they wait for user input before taking action. However, knowledge graphs require ongoing maintenance: new documents should be linked to relevant projects, repositories should be associated with their projects, orphaned tasks need homes, and stale items require attention. The autonomous orchestration layer addresses this need through continuous, proactive graph management.

\subsection{Architecture}

The autonomous agent operates on a configurable interval (default: 15 minutes) and follows an OODA-inspired loop:

\begin{enumerate}
    \item \textbf{Observe}: Query the graph for current state including recent changes, stale items, orphaned nodes, goal/project health metrics, connection opportunities, emails needing response, and meetings needing preparation
    \item \textbf{Orient}: Format observations into a structured context for LLM reasoning, including user writing samples and learned guidelines
    \item \textbf{Decide}: Use LLM to analyze context and select actions from available options, respecting current cognitive state
    \item \textbf{Act}: Execute selected actions and log results
\end{enumerate}

\subsection{State-Aware Decision Making}

The autonomous agent integrates with the Executive Function Layer to make state-appropriate decisions:

\begin{lstlisting}[language=Python]
async def _build_state_context(self) -> str:
    state = await state_estimator.get_current_state()
    in_recovery = await self._check_clinical_recovery()
    rules = ModeRules.get_rules(state.mode)

    return f"""
## Current User State
- Mode: {state.mode.value}
- Fatigue: {state.signals.fatigue_level:.0%}
- Hour of day: {datetime.now().hour}
- In clinical recovery: {in_recovery}

Based on mode '{state.mode.value}':
- Only suggest tasks with friction <= {rules.max_task_friction}
- For high-friction tasks, use SUGGEST_RESCHEDULE
- Peak energy hours: 8am-11am
"""
\end{lstlisting}

\subsection{Graph Observer}

The Graph Observer component provides comprehensive visibility into graph state through specialized queries:

\begin{lstlisting}[language=Python]
class GraphObserver:
    async def get_full_context(self) -> dict:
        return {
            "recent_changes": await self.get_recent_changes(),
            "stale_items": await self.get_stale_items(),
            "orphaned_nodes": await self.get_orphaned_nodes(),
            "goal_health": await self.get_goal_health(),
            "project_health": await self.get_project_health(),
            "pending_tasks": await self.get_pending_tasks(),
            "recent_documents": await self.get_recent_documents(),
            "connection_opportunities":
                await self.get_connection_opportunities(),
            "emails_needing_response":
                await self.get_actionable_emails(),
            "meetings_needing_prep":
                await self.get_pending_calendar_blocks(),
            "inbox_items": await self.get_firewall_inbox(),
        }
\end{lstlisting}

\paragraph{Connection Opportunities:} The observer identifies potential links based on:
\begin{itemize}
    \item \textbf{Name matching}: Documents/repositories containing project names (minimum 4 characters to avoid false positives)
    \item \textbf{Topic overlap}: Documents sharing topics with projects
    \item \textbf{Orphaned entities}: Tasks without projects, projects without goals
\end{itemize}

\subsection{Available Actions}

The autonomous agent can perform the following actions:

\paragraph{Digital Twin Actions:}
\begin{itemize}
    \item \texttt{DRAFT\_EMAIL}: Draft a reply to an incoming email in the user's voice, learning style from sent email samples
    \item \texttt{COMPILE\_CONTEXT\_PACK}: Prepare a briefing document for an upcoming meeting with relevant documents, tasks, and key points
    \item \texttt{SCHEDULE\_BLOCK}: Suggest focus time blocks for projects needing attention
    \item \texttt{PROCESS\_INBOX\_ITEM}: Process items captured by the interruption firewall during focus time
\end{itemize}

\paragraph{Graph Maintenance Actions:}
\begin{itemize}
    \item \texttt{LINK\_DOCUMENT}: Connect a document to a project based on name match or topic overlap
    \item \texttt{LINK\_REPOSITORY}: Connect a GitHub repository to a project
    \item \texttt{LINK\_TASK}: Connect an orphaned task to a project
    \item \texttt{LINK\_PROJECT\_TO\_GOAL}: Connect a project to a goal
    \item \texttt{CREATE\_TASK}: Create a task for a stalled project with no tasks
\end{itemize}

\paragraph{State-Aware Actions:}
\begin{itemize}
    \item \texttt{SUGGEST\_RESCHEDULE}: Suggest moving a high-friction task to a better time when user has more energy (e.g., peak morning hours)
    \item \texttt{FLAG\_FOR\_REVIEW}: Surface an item requiring human decision (triggers Discord notification); limited to 1 per cycle to avoid over-flagging
\end{itemize}

\subsection{Notification Deduplication}

The agent tracks recent notifications to avoid repetitive suggestions:

\begin{lstlisting}[language=Python]
## CRITICAL: Avoid Repetitive Notifications

Check "Recent Notification History" before any action:
- SCHEDULE_BLOCK: Don't re-suggest for same project in 24h
- COMPILE_CONTEXT_PACK: Don't re-compile for same meeting
- FLAG_FOR_REVIEW: Don't flag same issue repeatedly

Ask: "Would receiving this again annoy the user?"
\end{lstlisting}

\subsection{Externalized Prompts}

The autonomous agent's decision-making behavior is controlled by an externalized prompt stored as a markdown template:

\begin{lstlisting}[language=bash]
src/cognitex/prompts/autonomous_agent.md
\end{lstlisting}

This separation enables rapid iteration on agent behavior without code changes. The prompt template uses \texttt{\{variable\}} placeholders that are populated at runtime with current graph state, user writing samples, learned guidelines, and state context.

\subsection{Skip Tracking}

To prevent the agent from repeatedly suggesting the same actions after user dismissal, the system tracks actioned items:

\begin{itemize}
    \item Emails with any \texttt{EmailDraft} linked (regardless of draft status) are excluded from \texttt{get\_actionable\_emails()}
    \item Calendar events with any \texttt{ContextPack} linked are excluded from \texttt{get\_pending\_calendar\_blocks()}
    \item Projects with recent \texttt{SuggestedBlock} nodes are included in a skip list passed to the LLM prompt
    \item Recent notification history (48 hours) is provided to prevent duplicate notifications
\end{itemize}

\section{Executive Function Layer}
\label{sec:executive}

\subsection{Motivation}

Not all moments are equally suited for interruptions or demanding tasks. A user in deep focus should not be disturbed for routine notifications, while someone in fragmented attention mode may welcome task suggestions. After a cognitively exhausting clinical session, the user needs recovery time. The executive function layer models user cognitive state and manages both interruptions and task suggestions accordingly.

\subsection{Operating Modes}

The system defines six operating modes based on cognitive load theory:

\begin{table}[t]
\centering
\caption{Operating modes and their characteristics.}
\label{tab:operating_modes}
\begin{tabular}{lp{5cm}ll}
\toprule
\textbf{Mode} & \textbf{Description} & \textbf{Max Friction} & \textbf{Interruption} \\
\midrule
DEEP\_FOCUS & Sustained cognitive work, long blocks & 5 & Minimal \\
FRAGMENTED & Context switching, short blocks & 3 & Moderate \\
OVERLOADED & Recovery mode, maintenance only & 1 & Critical only \\
AVOIDANT & Avoiding high-friction tasks & 2 & Low priority \\
HYPERFOCUS & Intense focus, may miss signals & 5 & Emergency only \\
TRANSITION & Between modes, ramping up/down & 3 & Batched \\
\bottomrule
\end{tabular}
\end{table}

Each mode defines specific rules:

\begin{lstlisting}[language=Python]
@dataclass
class ModeRules:
    mode: OperatingMode
    allowed_task_types: list[str]
    max_task_friction: int  # 1-5 scale
    min_block_minutes: int | None
    notification_gate: str  # all, batched, critical_only
    interrupt_for: list[str]  # urgent, family, emergency
    ui_density: str  # full, compact, minimal
    auto_actions: list[str]  # allowed auto-actions
\end{lstlisting}

\subsection{Continuous Signals}

Beyond discrete modes, the system tracks continuous signals:

\begin{lstlisting}[language=Python]
@dataclass
class ContinuousSignals:
    # Time constraints
    available_block_minutes: int | None
    time_to_next_commitment_minutes: int | None

    # Load indicators
    interruption_pressure: float  # 0-1
    fatigue_level: float          # 0-1
    fatigue_slope: float          # -1 to +1

    # Attention state
    focus_score: float | None     # 0-1

    # Context
    location: str | None          # home, office, travel
    device: str | None            # desktop, mobile
\end{lstlisting}

\subsection{Temporal Energy Model}

The system includes a diurnal energy model that captures typical energy patterns throughout the day:

\begin{lstlisting}[language=Python]
class TemporalEnergyModel:
    # Diurnal pattern: peak 8-11am, decline through day
    DEFAULT_CURVE: dict[int, float] = {
        0: 0.15, 1: 0.1, 2: 0.1, 3: 0.1,
        4: 0.15, 5: 0.25, 6: 0.45, 7: 0.7,
        8: 0.9,   # entering peak zone
        9: 0.95,  # peak
        10: 0.95, # peak
        11: 0.9,  # still high
        12: 0.75, # lunch dip
        13: 0.65, 14: 0.55, 15: 0.5,
        16: 0.45, 17: 0.4, 18: 0.35,
        19: 0.3, 20: 0.25, 21: 0.2,
        22: 0.15, 23: 0.15,
    }

    async def get_expected_energy(self, hour: int) -> float:
        """Get expected energy, checking learned overrides."""
        learned = await self._get_learned_pattern(hour)
        return learned if learned else self.DEFAULT_CURVE.get(hour, 0.5)
\end{lstlisting}

The model supports learning from observations: when tasks are completed or deferred, the outcome is recorded with the hour and state context, allowing the curve to adapt to the user's actual patterns.

\subsection{Clinical Session Detection}

High-drain events (clinical sessions, board meetings, major presentations) are identified from calendar events:

\begin{lstlisting}[language=Python]
class EventDrainLevel(str, Enum):
    CLINICAL = "clinical"   # Rest of day recovery
    HIGH = "high"           # 1hr recovery buffer
    MEDIUM = "medium"       # Normal meetings
    LOW = "low"             # 1:1s, syncs
    RECOVERY = "recovery"   # Blocked recovery time

def classify_event_drain(event: dict) -> EventDrainLevel:
    title = event.get("summary", "").upper()

    if "[CLINICAL]" in title:
        return EventDrainLevel.CLINICAL

    if any(kw in title.lower() for kw in
           ["board", "presentation", "interview"]):
        return EventDrainLevel.HIGH

    if any(kw in title.lower() for kw in
           ["1:1", "sync", "catch up"]):
        return EventDrainLevel.LOW

    return EventDrainLevel.MEDIUM
\end{lstlisting}

When a clinical session ends, the system automatically:
\begin{enumerate}
    \item Sets operating mode to OVERLOADED for the rest of the day
    \item Stores recovery expiry in Redis for persistence
    \item Sends a notification explaining the mode change
    \item Filters all subsequent task suggestions to low-friction only
\end{enumerate}

\subsection{State Estimation}

The State Estimator infers current operating mode from multiple signals:

\begin{lstlisting}[language=Python]
def _infer_mode(
    self,
    signals: ContinuousSignals,
    recent_tasks: list[dict] | None = None,
) -> OperatingMode:
    # Check for overload
    if (signals.fatigue_level > 0.8 and
        signals.interruption_pressure > 0.7):
        return OperatingMode.OVERLOADED

    # Check for avoidance (repeated deferrals)
    if recent_tasks:
        deferral_rate = self._calculate_deferral_rate(recent_tasks)
        if deferral_rate > 0.5:
            return OperatingMode.AVOIDANT

    # Check for deep focus potential
    if (signals.available_block_minutes and
        signals.available_block_minutes >= 90 and
        signals.interruption_pressure < 0.3):
        return OperatingMode.DEEP_FOCUS

    return OperatingMode.FRAGMENTED
\end{lstlisting}

\subsection{Interruption Firewall}

The Interruption Firewall filters incoming notifications based on current mode:

\begin{lstlisting}[language=Python]
class InterruptionFirewall:
    async def should_interrupt(
        self,
        urgency: str,  # low, normal, high, urgent
        source: str    # email, calendar, agent, etc.
    ) -> tuple[bool, str]:
        """Returns (should_interrupt, reason)"""
        mode = await self.get_current_mode()
        rules = ModeRules.get_rules(mode)

        if urgency == "urgent":
            return True, "Urgent items always interrupt"
        if mode == OperatingMode.DEEP_FOCUS:
            return False, "Deep work mode - queued"
        if mode == OperatingMode.OVERLOADED:
            if urgency not in ["high", "urgent"]:
                return False, "Recovery mode - low priority queued"
        # ... mode-specific logic

    async def capture_for_later(self, item: dict) -> None:
        """Queue item in firewall inbox for later processing."""
        await self._store_inbox_item(item)
\end{lstlisting}

Items blocked by the firewall are queued in the ``firewall inbox'' and surfaced during appropriate moments (mode transitions, scheduled review times) via the \texttt{PROCESS\_INBOX\_ITEM} action.

\subsection{State Snapshots}

State snapshots are persisted to Neo4j for historical analysis and pattern detection:

\begin{lstlisting}[language=SQL]
CREATE (ss:StateSnapshot {
    id: $snapshot_id,
    mode: $mode,
    fatigue_level: $fatigue_level,
    interruption_pressure: $interruption_pressure,
    focus_score: $focus_score,
    context_notes: $notes,
    created_at: datetime()
})
\end{lstlisting}

\section{Behavioral Learning System}
\label{sec:learning}

\subsection{Motivation}

A personal agent that never learns from user feedback will repeatedly make the same mistakes. The behavioral learning system addresses this by extracting patterns from user interactions, tracking proposal acceptance rates, calibrating duration estimates, predicting deferral risk, and feeding learned insights back into agent prompts.

\subsection{Architecture}

The learning system comprises four interconnected components:

\begin{enumerate}
    \item \textbf{Proposal Learning}: Tracks which task proposals are accepted/rejected by priority, project, and context
    \item \textbf{Duration Calibration}: Compares estimated vs. actual task durations to build adjustment factors
    \item \textbf{Deferral Prediction}: Identifies tasks at risk of procrastination based on historical patterns
    \item \textbf{Rule Extraction}: Converts observed patterns into explicit preference rules
\end{enumerate}

\subsection{Preference Rules}

Preference rules encode learned user behaviors with a lifecycle model:

\begin{lstlisting}[language=Python]
@dataclass
class PreferenceRule:
    rule_name: str
    rule_type: str  # scheduling, notification, proposal, etc.
    conditions: dict  # When rule applies
    action: dict  # What to do
    confidence: float  # 0-1
    lifecycle: str  # candidate, validated, deprecated
    applications: int  # Times applied
    successful_applications: int
    success_rate: float | None
\end{lstlisting}

Rules progress through lifecycle stages:
\begin{itemize}
    \item \textbf{Candidate}: Newly extracted from observations, minimum 3 samples, confidence $\geq$ 0.6
    \item \textbf{Validated}: Applied 5+ times with success rate $\geq$ 0.7
    \item \textbf{Deprecated}: Success rate dropped below 0.5 for 10+ applications, or superseded by better rule
\end{itemize}

\subsection{Proposal Pattern Learning}

The system tracks task proposal outcomes by multiple dimensions:

\begin{lstlisting}[language=SQL]
CREATE TABLE task_proposals (
    id TEXT PRIMARY KEY,
    title TEXT NOT NULL,
    project_id TEXT,
    priority TEXT,
    source TEXT,  -- agent, trigger
    status TEXT,  -- pending, approved, rejected, expired
    created_at TIMESTAMP DEFAULT NOW(),
    decided_at TIMESTAMP,
    user_feedback TEXT
);
\end{lstlisting}

Patterns extracted include:
\begin{itemize}
    \item Approval rate by priority level (high, medium, low)
    \item Approval rate by project
    \item Approval rate by time of day
    \item Common rejection reasons
\end{itemize}

\subsection{Duration Calibration}

For tasks with estimates, the system tracks actual completion times:

\begin{lstlisting}[language=Python]
async def get_calibration_summary(project_id: str = None):
    """Get duration calibration data."""
    # Compare estimated_minutes vs actual_minutes
    # Calculate pace_factor = actual / estimated
    # Aggregate by project and overall
    return {
        "samples": count,
        "avg_error": (avg_actual - avg_estimated) / avg_estimated,
        "by_project": {...}
    }
\end{lstlisting}

This enables personalized duration adjustments: ``You typically take 30\% longer than estimated on Project X.''

\subsection{Deferral Risk Prediction}

The system identifies tasks at risk of indefinite deferral:

\begin{lstlisting}[language=Python]
async def get_high_risk_tasks(min_risk: float = 0.5) -> list[dict]:
    """Get tasks with high deferral risk."""
    # Risk factors:
    # - High deferral_count
    # - Long time since creation with no progress
    # - High friction + current low energy
    # - Similar tasks previously abandoned
    return [
        {
            "task_id": ...,
            "title": ...,
            "risk_score": 0.85,
            "factors": ["deferred 3x", "high friction", "no MVS"],
            "recommended_intervention": "Break into smaller steps"
        }
    ]
\end{lstlisting}

\subsection{State Observation Recording}

Task outcomes are recorded with full state context for temporal pattern learning:

\begin{lstlisting}[language=SQL]
CREATE TABLE state_observations (
    id SERIAL PRIMARY KEY,
    task_id TEXT,
    task_title TEXT,
    outcome TEXT,  -- completed, deferred, abandoned
    mode TEXT,
    fatigue_level FLOAT,
    hour_of_day INTEGER,
    day_of_week INTEGER,
    post_clinical BOOLEAN DEFAULT FALSE,
    energy_cost TEXT,  -- high, medium, low
    task_friction INTEGER,
    observed_at TIMESTAMP DEFAULT NOW()
);
\end{lstlisting}

This enables learning patterns like: ``High-friction tasks are rarely completed after 3pm'' or ``Post-clinical deferrals are 3x higher.''

\subsection{Learned Guidelines Injection}

Extracted patterns are injected into the autonomous agent prompt:

\begin{lstlisting}[language=markdown]
## LEARNED GUIDELINES & FEEDBACK
Apply these lessons from previous interactions:

### Validated Rules
- [SCHEDULING] Morning hours (8-11am) have 85% task completion
  vs 45% afternoon (confidence: 0.89, 47 samples)
- [PROPOSAL] High-priority tasks for "Research" project have
  92% approval (confidence: 0.91, 24 samples)

### Recent Corrections
- User rejected "Schedule focus block for Project X" with
  feedback: "Already have enough focus time scheduled"
- Avoid re-suggesting similar blocks for 48 hours
\end{lstlisting}

\subsection{Learning Dashboard}

The web interface provides visibility into learning system state:

\begin{itemize}
    \item \textbf{Proposal Patterns}: Approval rates by priority and project
    \item \textbf{Duration Calibration}: Estimate accuracy with adjustment recommendations
    \item \textbf{Deferral Risk}: High-risk tasks with intervention suggestions
    \item \textbf{Learned Patterns}: Extracted temporal and behavioral patterns
    \item \textbf{Preference Rules}: Rules by lifecycle stage with confidence scores
    \item \textbf{Policy Updates}: Last learning cycle results
\end{itemize}

\section{Semantic Graph Construction}
\label{sec:graph}

\subsection{Entity Extraction Pipeline}

The ingestion pipeline transforms raw data from external sources into graph entities. \Cref{alg:email} presents the email processing algorithm.

\begin{algorithm}[t]
\caption{Email Ingestion Pipeline}
\label{alg:email}
\begin{algorithmic}[1]
\Procedure{IngestEmail}{$email\_data$}
    \State Create sender Person node
    \For{each recipient in to $\cup$ cc}
        \State Create recipient Person node
    \EndFor
    \State $classification \gets$ \Call{LLMClassify}{$email\_data$}
    \State Create Email node with classification
    \State Create SENT\_BY relationship
    \For{each recipient}
        \State Create RECEIVED\_BY relationship
    \EndFor
    \State \Call{InferWorksWithRelationships}{sender, recipients}
    \If{$email\_data.action\_required$}
        \State \Call{InferTasks}{$email\_data$}
    \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

\subsection{Email Classification}

The LLM classifies emails into four categories:
\begin{itemize}
    \item \textbf{actionable}: Requires user response or action
    \item \textbf{delegable}: Could be handled by someone else
    \item \textbf{fyi}: Informational only
    \item \textbf{spam}: Unwanted/promotional
\end{itemize}

Additional properties extracted include urgency (low, medium, high), action\_required (boolean), and suggested\_action.

\subsection{Calendar Event Classification}

Calendar events are classified by cognitive drain level during sync:

\begin{lstlisting}[language=SQL]
MERGE (e:Event {gcal_id: $gcal_id})
SET e.title = $title,
    e.start = datetime($start),
    e.end = datetime($end),
    e.drain_level = $drain_level
\end{lstlisting}

Events with \texttt{drain\_level = 'clinical'} trigger special handling by the executive function layer.

\subsection{Relationship Inference}

Beyond explicit relationships, the system infers latent connections. The \texttt{WORKS\_WITH} relationship is computed from email interactions:

\begin{lstlisting}[language=SQL]
MATCH (e:Email)-[:SENT_BY]->(sender:Person)
MATCH (e)-[:RECEIVED_BY]->(recipient:Person)
WHERE sender <> recipient
MERGE (sender)-[r:WORKS_WITH]->(recipient)
ON CREATE SET r.interaction_count = 1
ON MATCH SET r.interaction_count = r.interaction_count + 1
\end{lstlisting}

\subsection{Autonomous Link Creation}

The autonomous agent creates links based on name matching:

\begin{lstlisting}[language=SQL]
-- Documents matching project names (4+ char minimum)
MATCH (d:Document), (p:Project)
WHERE NOT (d)-[:BELONGS_TO|REFERENCES]->(p)
  AND (
    (size(p.title) >= 4 AND
     toLower(d.name) CONTAINS toLower(p.title))
    OR (size(d.name) >= 4 AND
        toLower(p.title) CONTAINS toLower(d.name))
  )
RETURN d.drive_id, d.name, p.id, p.title
\end{lstlisting}

Similar queries identify repository-project matches, task-project matches, and project-goal matches.

\section{Document Chunking and Embedding Pipeline}
\label{sec:chunking}

\subsection{Chunking Strategy}

Documents are split into overlapping chunks optimized for embedding model context windows. The configuration uses:

\begin{itemize}
    \item Chunk size: 1,200 characters ($\approx$300 tokens)
    \item Overlap: 200 characters ($\approx$50 tokens)
    \item Minimum chunk size: 100 characters
\end{itemize}

The chunking algorithm preserves semantic boundaries by splitting on paragraph boundaries first, then sentence boundaries for oversized paragraphs.

\subsection{Embedding Generation}

Chunks are embedded using BAAI/bge-base-en-v1.5, producing 768-dimensional vectors stored in PostgreSQL with pgvector.

\subsection{Semantic Search}

Search combines vector similarity with metadata filtering:

\begin{lstlisting}[language=SQL]
SELECT e.entity_id, dc.content, d.name,
       1 - (e.embedding <=> $1::vector) as similarity
FROM embeddings e
JOIN document_chunks dc ON dc.drive_id || ':' ||
     dc.chunk_index = e.entity_id
JOIN documents d ON d.drive_id = dc.drive_id
WHERE e.entity_type = 'chunk'
ORDER BY e.embedding <=> $1::vector
LIMIT $2
\end{lstlisting}

\subsection{Auto-Indexing Pipeline}

When Drive files change (detected via polling), the system automatically re-indexes by extracting content, deleting old chunks and embeddings, generating new chunks with embeddings, and analyzing chunks for graph integration.

\section{Trigger and Notification System}
\label{sec:triggers}

\subsection{Trigger Types}

The system supports four trigger categories:

\paragraph{Scheduled Triggers:} APScheduler provides cron-based scheduling for:
\begin{itemize}
    \item Morning briefings (8:00 AM)
    \item Evening reviews (6:00 PM)
    \item Hourly monitoring (9 AM--6 PM)
    \item Overdue task checks (9:30 AM)
    \item GitHub sync (every 4 hours)
    \item Drive polling (every 15 minutes)
    \item Stuck task detection (every 30 minutes)
\end{itemize}

\paragraph{Event-Driven Triggers:} Multiple pub/sub mechanisms enable reactive processing:
\begin{itemize}
    \item \textbf{Gmail}: Google Cloud Pub/Sub pull subscription (\texttt{\_gmail\_pubsub\_listener}) receives real-time notifications when new emails arrive
    \item \textbf{Internal events}: Redis pub/sub handles calendar changes, task updates, Drive modifications, and calendar event completions (for clinical session detection)
\end{itemize}

\paragraph{Autonomous Triggers:} The autonomous agent runs on its own schedule (default: every 15 minutes) for proactive graph maintenance and Digital Twin operations.

\paragraph{User-Initiated Triggers:} Direct interaction via CLI, Discord, web dashboard, or REST API.

\subsection{Clinical Session Handling}

When a calendar event ends, the system checks for clinical sessions:

\begin{lstlisting}[language=Python]
async def _adjust_state_after_event(self, event: dict):
    drain = classify_event_drain(event)

    if drain == EventDrainLevel.CLINICAL:
        # Set OVERLOADED mode for rest of day
        await state_estimator.update_state(
            mode=OperatingMode.OVERLOADED,
            fatigue=0.95,
            notes=f"Post-clinical recovery: {event.get('summary')}"
        )

        # Persist in Redis for service restart survival
        await redis.set(
            "cognitex:clinical_recovery_until",
            datetime.now().replace(hour=23, minute=59).isoformat(),
            ex=86400
        )

        # Notify user
        await self._send_notification(
            "**Clinical session ended**\n\n"
            "Switching to recovery mode. "
            "Only low-energy tasks will be suggested.",
            urgency="low"
        )
\end{lstlisting}

\subsection{Stuck Task Detection}

The system monitors for tasks where the user has been working significantly longer than estimated:

\begin{lstlisting}[language=Python]
async def _check_stuck_tasks(self):
    """Check for tasks at 150%+ of estimated time."""
    # Find in-progress tasks with estimates
    # Calculate elapsed time since started_at
    # If elapsed > 1.5 * estimated_minutes:
    #   Send whisper notification offering to break down task
\end{lstlisting}

\subsection{Approval Workflow}

High-risk actions follow a staged approval process:

\begin{enumerate}
    \item Agent proposes action (e.g., send email)
    \item Action staged in Redis with metadata
    \item Notification published to Discord with approve/reject buttons
    \item User reviews and approves/edits/rejects
    \item On approval: execute action, record decision trace
    \item On rejection: store feedback for learning
\end{enumerate}

\subsection{Notification Routing}

Notifications are routed through the Interruption Firewall before delivery:

\begin{enumerate}
    \item Notification generated (email analysis, agent flag, etc.)
    \item Firewall evaluates against current operating mode
    \item If allowed: publish to Redis pub/sub for Discord delivery
    \item If blocked: queue in firewall inbox for later processing
\end{enumerate}

\section{Web Dashboard}
\label{sec:web}

\subsection{Overview}

The web dashboard provides a comprehensive interface for system interaction and monitoring, built with FastAPI and Jinja2 templates with HTMX for dynamic updates.

\subsection{Pages}

\paragraph{Dashboard (Home):} Overview of system state including pending tasks, upcoming events, recent emails requiring action, and quick stats.

\paragraph{Tasks:} Task management with inline editing, status updates, energy cost assignment, due date setting, duration estimates, and project/goal linking.

\paragraph{Projects:} Project management with task rollup, progress tracking, and goal association.

\paragraph{Goals:} Strategic goal management with project linkage and progress visualization.

\paragraph{Documents:} Document browser with semantic search, topic filtering, and project linking.

\paragraph{Graph:} Interactive graph visualization using vis.js with node type filtering and dynamic relationship display.

\paragraph{Agent Log:} Autonomous agent action history with:
\begin{itemize}
    \item Action timeline (link\_document, draft\_email, suggest\_reschedule, etc.)
    \item Filtering by status (completed, failed, pending) and source (agent, trigger, user)
    \item Statistics: total actions, last 24h, failed count
    \item Manual ``Run Cycle Now'' button for testing
\end{itemize}

\paragraph{Twin (Digital Twin Review):} Review and approve agent-generated outputs:
\begin{itemize}
    \item \textbf{Email Drafts}: Pending drafts with approve/edit/discard actions
    \item \textbf{Context Packs}: Meeting briefings with key points and linked documents
    \item \textbf{Suggested Blocks}: Focus time suggestions with approve/dismiss
\end{itemize}

\paragraph{State:} Operating mode management with:
\begin{itemize}
    \item Current mode display with description
    \item Mode switching controls
    \item Continuous signal adjustment (fatigue, interruption pressure)
    \item Mode rules display (allowed task types, max friction, notification gate)
    \item Today's switch statistics
    \item Firewall inbox for captured items
\end{itemize}

\paragraph{Learning:} Learning system insights with:
\begin{itemize}
    \item Proposal pattern statistics by priority and project
    \item Duration calibration with accuracy metrics
    \item High-risk deferral tasks with intervention suggestions
    \item Learned patterns with confidence scores
    \item Preference rules by lifecycle stage
    \item Policy update status and manual trigger
\end{itemize}

\paragraph{Settings:} Configuration management for autonomous agent interval, integration settings, remote session sync configuration with API key display and setup instructions, and user preferences.

\paragraph{Session Linking:} Human-in-the-loop interface for linking coding sessions to projects:
\begin{itemize}
    \item Unlinked sessions with fuzzy match suggestions
    \item Confidence scores based on path-to-title similarity
    \item One-click approval for high-confidence matches
    \item Bulk ``Accept All High-Confidence'' action for 90\%+ matches
    \item Manual project selection dropdown for low-confidence or unmatched sessions
    \item Skip option for sessions without relevant projects
\end{itemize}

\section{Evaluation}
\label{sec:evaluation}

\subsection{System Metrics}

The current deployment demonstrates the following characteristics:

\paragraph{Data Scale:}
\begin{itemize}
    \item 153,000+ documents indexed from Google Drive
    \item 3,400+ document chunks with embeddings
    \item 50+ documents with deep semantic analysis
    \item 5 GitHub repositories indexed
    \item 100+ coding sessions from multiple machines (local server + remote laptop)
    \item 102 sessions linked to 4 projects via fuzzy matching
    \item $\approx$15,000 lines of application code
\end{itemize}

\paragraph{Autonomous Agent Activity:}
\begin{itemize}
    \item Average 5 actions per 15-minute cycle
    \item Action types: link\_document (most common), link\_repository, link\_project\_to\_goal, create\_task, draft\_email, compile\_context\_pack, suggest\_reschedule
    \item Near-zero flag\_for\_review rate with optimized prompt
    \item State-aware suggestions: high-friction tasks automatically rescheduled to peak hours
\end{itemize}

\paragraph{Learning System:}
\begin{itemize}
    \item 50+ preference rules extracted
    \item Duration calibration from 100+ completed tasks
    \item Proposal approval rate: 75\% (improving with learned guidelines)
    \item Temporal patterns: clear peak productivity 8--11am confirmed
\end{itemize}

\paragraph{Performance:}
\begin{itemize}
    \item Embedding generation: $\approx$0.5s per chunk
    \item Semantic search: $<$100ms for top-10 results
    \item Graph queries: $<$50ms for typical traversals
    \item Agent response: 3--15s depending on tool usage
    \item Autonomous cycle: 5--10s total
    \item State estimation: $<$10ms
\end{itemize}

\subsection{Qualitative Assessment}

The system successfully demonstrates:
\begin{enumerate}
    \item \textbf{Context Persistence}: The agent maintains awareness of email threads, contacts, and commitments across sessions
    \item \textbf{Relationship Inference}: WORKS\_WITH relationships accurately reflect collaboration patterns
    \item \textbf{Risk Stratification}: Approval workflow prevents unintended actions while allowing autonomous task creation
    \item \textbf{Semantic Search}: Document retrieval surfaces relevant content across diverse file types
    \item \textbf{Proactive Linking}: Autonomous agent successfully identifies and creates document-project, repository-project, and project-goal links
    \item \textbf{State Awareness}: Agent respects cognitive mode, avoiding demanding suggestions during recovery
    \item \textbf{Clinical Recovery}: Post-clinical OVERLOADED mode successfully prevents inappropriate task suggestions
    \item \textbf{Behavioral Learning}: Preference rules improve suggestion quality over time
    \item \textbf{Temporal Adaptation}: Peak hour suggestions align with learned productivity patterns
\end{enumerate}

\subsection{Limitations}

Current limitations include: (1) decision traces captured but not yet used for model fine-tuning, (2) Calendar uses polling rather than push notifications, (3) limited automated testing, (4) single-user deployment only, and (5) clinical detection requires manual \texttt{[CLINICAL]} tagging rather than automatic inference from calendar patterns.

\section{Discussion and Future Work}
\label{sec:discussion}

\subsection{Design Decisions}

Several architectural decisions warrant discussion:

\paragraph{Hybrid Storage:} The combination of Neo4j, PostgreSQL, and Redis adds operational complexity but enables optimal data models for each use case. Graph traversal, vector similarity, and low-latency caching each have distinct requirements poorly served by a single database.

\paragraph{Risk Stratification:} The three-tier risk model represents a pragmatic balance between autonomy and control. Future work may explore dynamic risk assessment based on context and learned user preferences.

\paragraph{Autonomous Orchestration:} The decision to implement proactive graph management as a separate always-running agent (rather than integrating into reactive flows) provides clear separation of concerns and predictable behavior.

\paragraph{State Modeling:} Six operating modes provide sufficient granularity for meaningful behavior differentiation without overwhelming complexity. The continuous signals complement discrete modes for nuanced decision-making.

\paragraph{Learning System:} The rule lifecycle model (candidate $\rightarrow$ validated $\rightarrow$ deprecated) prevents stale rules from degrading system performance while allowing rapid adaptation to changing preferences.

\paragraph{Externalized Prompts:} Storing prompts as markdown files rather than embedding in code enables rapid iteration on agent behavior. This is particularly valuable for the autonomous agent where prompt engineering significantly affects suggestion quality.

\subsection{Future Directions}

Several extensions are planned:
\begin{enumerate}
    \item Behavioral fine-tuning using captured decision traces (infrastructure ready---traces captured with training flags---but model fine-tuning not yet implemented)
    \item Calendar push notifications via Google Cloud Pub/Sub (Gmail already uses Pub/Sub; Calendar still uses polling)
    \item Multi-modal input processing (images, attachments)
    \item Collaborative features for team environments
    \item Granular privacy controls
    \item Physiological signal integration (wearables) for state estimation
    \item Automatic clinical session detection from calendar patterns (currently requires manual \texttt{[CLINICAL]} tagging)
    \item Cross-entity semantic linking (email content to documents via embedding similarity)
\end{enumerate}

\paragraph{Already Implemented (moved from future work):}
\begin{itemize}
    \item \textbf{Gmail push notifications}: Pull-based Pub/Sub subscription via \texttt{\_gmail\_pubsub\_listener()} receives real-time notifications from Gmail
    \item \textbf{Agent self-improvement through learned guidelines}: Preference rules are extracted from user feedback and injected into autonomous agent prompts via the \texttt{\{learned\_guidelines\}} template variable
    \item \textbf{Personalized temporal energy curves}: The \texttt{TemporalEnergyModel} updates learned overrides via \texttt{update\_from\_observation()} as task outcomes are recorded with hour-of-day context
\end{itemize}

\subsection{Ethical Considerations}

Personal agent systems raise important ethical questions around data privacy (broad access to personal communications), autonomy versus control (approval friction versus safety), transparency (decision trace accessibility), and dependency (reduced cognitive engagement from over-reliance).

\section{Conclusion}
\label{sec:conclusion}

\cognitex{} demonstrates a viable architecture for personal AI agents that balance autonomous capability with human oversight. The hybrid knowledge representation---combining property graphs, vector embeddings, and session caching---provides a robust foundation for personal information management. The risk-stratified tool system enables meaningful automation while preserving user agency for consequential actions.

Key contributions include the autonomous graph orchestration layer for proactive graph maintenance, the executive function layer for cognitive state awareness with six operating modes and temporal energy modeling, the behavioral learning system for preference extraction and guideline injection, clinical session detection for automatic recovery mode, and the comprehensive web dashboard for system visibility.

The agent's ability to respect user cognitive state---avoiding demanding suggestions when fatigued, automatically entering recovery mode after clinical sessions, and learning which proposals are accepted---represents a significant advance toward truly adaptive personal AI. The learning system ensures the agent improves over time, extracting patterns from user feedback and injecting them back into decision-making.

While significant work remains---particularly in leveraging captured traces for model fine-tuning and expanding physiological signal integration---the current implementation provides a functional reference architecture for trustworthy personal AI systems. We release the complete codebase to support further research in this domain.

\section*{Acknowledgments}

This work was developed independently with assistance from Claude (Anthropic) for code generation and architectural refinement.

\bibliographystyle{plainnat}
\begin{thebibliography}{10}

\bibitem[Dong et al.(2014)]{dong2014knowledge}
Dong, X., Gabrilovich, E., Heitz, G., Horn, W., Lao, N., Murphy, K., Strohmann, T., Sun, S., \& Zhang, W. (2014).
\newblock Knowledge vault: A web-scale approach to probabilistic knowledge fusion.
\newblock In \emph{Proceedings of the 20th ACM SIGKDD International Conference on Knowledge Discovery and Data Mining} (pp. 601--610).

\bibitem[Horvitz(1999)]{horvitz1999principles}
Horvitz, E. (1999).
\newblock Principles of mixed-initiative user interfaces.
\newblock In \emph{Proceedings of the SIGCHI Conference on Human Factors in Computing Systems} (pp. 159--166).

\bibitem[Karpukhin et al.(2020)]{karpukhin2020dense}
Karpukhin, V., Oguz, B., Min, S., Lewis, P., Wu, L., Edunov, S., Chen, D., \& Yih, W.-T. (2020).
\newblock Dense passage retrieval for open-domain question answering.
\newblock In \emph{Proceedings of the 2020 Conference on Empirical Methods in Natural Language Processing} (pp. 6769--6781).

\bibitem[McKinsey(2012)]{mckinsey2012social}
McKinsey Global Institute. (2012).
\newblock The social economy: Unlocking value and productivity through social technologies.
\newblock \emph{McKinsey \& Company}.

\bibitem[Nickel et al.(2016)]{nickel2016review}
Nickel, M., Murphy, K., Tresp, V., \& Gabrilovich, E. (2016).
\newblock A review of relational machine learning for knowledge graphs.
\newblock \emph{Proceedings of the IEEE}, 104(1), 11--33.

\bibitem[Park et al.(2023)]{park2023generative}
Park, J.~S., O'Brien, J.~C., Cai, C.~J., Morris, M.~R., Liang, P., \& Bernstein, M.~S. (2023).
\newblock Generative agents: Interactive simulacra of human behavior.
\newblock In \emph{Proceedings of the 36th Annual ACM Symposium on User Interface Software and Technology} (pp. 1--22).

\bibitem[Qin et al.(2023)]{qin2023toolllm}
Qin, Y., Liang, S., Ye, Y., Zhu, K., Yan, L., Lu, Y., Lin, Y., Cong, X., Tang, X., Qian, B., et al. (2023).
\newblock ToolLLM: Facilitating large language models to master 16000+ real-world APIs.
\newblock \emph{arXiv preprint arXiv:2307.16789}.

\bibitem[Sauermann et al.(2005)]{sauermann2005semantic}
Sauermann, L., Bernardi, A., \& Dengel, A. (2005).
\newblock Overview and outlook on the semantic desktop.
\newblock In \emph{Proceedings of the ISWC 2005 Workshop on the Semantic Desktop}.

\bibitem[Shinn et al.(2023)]{shinn2023reflexion}
Shinn, N., Cassano, F., Gopinath, A., Narasimhan, K., \& Yao, S. (2023).
\newblock Reflexion: Language agents with verbal reinforcement learning.
\newblock \emph{Advances in Neural Information Processing Systems}, 36.

\bibitem[Yao et al.(2022)]{yao2022react}
Yao, S., Zhao, J., Yu, D., Du, N., Shafran, I., Narasimhan, K., \& Cao, Y. (2022).
\newblock ReAct: Synergizing reasoning and acting in language models.
\newblock \emph{arXiv preprint arXiv:2210.03629}.

\bibitem[Yorke-Smith et al.(2012)]{yorke2012proactive}
Yorke-Smith, N., Saadati, S., Myers, K.~L., \& Morley, D.~N. (2012).
\newblock The design of a proactive personal agent for task management.
\newblock \emph{International Journal on Artificial Intelligence Tools}, 21(01), 1250004.

\end{thebibliography}

\appendix

\section{Configuration Reference}
\label{app:config}

\begin{lstlisting}[language=bash,caption={Environment configuration variables}]
# Database URLs
DATABASE_URL=postgresql://cognitex:pass@localhost:5432/cognitex
NEO4J_URI=bolt://localhost:7687
NEO4J_PASSWORD=<secret>
REDIS_URL=redis://localhost:6379/0

# LLM Provider
TOGETHER_API_KEY=<secret>
TOGETHER_MODEL_PLANNER=deepseek-ai/DeepSeek-V3
TOGETHER_MODEL_EXECUTOR=deepseek-ai/DeepSeek-V3
TOGETHER_MODEL_EMBEDDING=BAAI/bge-base-en-v1.5

# Google APIs
GOOGLE_CLIENT_ID=<client_id>
GOOGLE_CLIENT_SECRET=<secret>

# Discord
DISCORD_BOT_TOKEN=<secret>
DISCORD_CHANNEL_ID=<channel_id>
DISCORD_USER_ID=<user_id>

# GitHub
GITHUB_TOKEN=<secret>

# Autonomous Agent
AUTONOMOUS_AGENT_ENABLED=true
AUTONOMOUS_AGENT_INTERVAL_MINUTES=15
\end{lstlisting}

\section{CLI Command Reference}
\label{app:cli}

\begin{lstlisting}[language=bash,caption={Primary CLI commands}]
# System
cognitex status              # Configuration and memory stats
cognitex auth                # OAuth2 authentication flow
cognitex graph               # Neo4j statistics
cognitex web                 # Start web dashboard

# Data Synchronization
cognitex sync                # Incremental Gmail sync
cognitex calendar            # Google Calendar sync
cognitex drive-sync          # Drive metadata sync
cognitex deep-index          # Document chunking and embedding

# Queries
cognitex tasks               # List tasks
cognitex contacts            # List people
cognitex doc-search "query"  # Semantic search

# Coding Sessions
cognitex sessions-sync       # Sync local Claude Code sessions
cognitex sessions-context    # View project development context

# Agent
cognitex agent-chat          # Interactive mode
cognitex briefing            # Morning briefing
cognitex approvals           # Pending approvals

# State Management
cognitex state               # View current state
cognitex state set-mode MODE # Set operating mode

# Learning
cognitex learning summary    # Learning system stats
cognitex learning patterns   # View learned patterns
\end{lstlisting}

\section{Neo4j Schema}
\label{app:schema}

\begin{lstlisting}[language=SQL,caption={Neo4j constraints and indexes}]
// Core Constraints
CREATE CONSTRAINT person_email_unique
  FOR (p:Person) REQUIRE p.email IS UNIQUE;
CREATE CONSTRAINT email_gmail_id_unique
  FOR (e:Email) REQUIRE e.gmail_id IS UNIQUE;
CREATE CONSTRAINT event_gcal_id_unique
  FOR (ev:Event) REQUIRE ev.gcal_id IS UNIQUE;
CREATE CONSTRAINT task_id_unique
  FOR (t:Task) REQUIRE t.id IS UNIQUE;
CREATE CONSTRAINT document_drive_id_unique
  FOR (d:Document) REQUIRE d.drive_id IS UNIQUE;
CREATE CONSTRAINT project_id_unique
  FOR (p:Project) REQUIRE p.id IS UNIQUE;
CREATE CONSTRAINT goal_id_unique
  FOR (g:Goal) REQUIRE g.id IS UNIQUE;
CREATE CONSTRAINT state_snapshot_id_unique
  FOR (ss:StateSnapshot) REQUIRE ss.id IS UNIQUE;
CREATE CONSTRAINT coding_session_id_unique
  FOR (cs:CodingSession) REQUIRE cs.session_id IS UNIQUE;

// State and Learning Indexes
CREATE INDEX state_snapshot_mode
  FOR (ss:StateSnapshot) ON (ss.mode);
CREATE INDEX state_snapshot_created
  FOR (ss:StateSnapshot) ON (ss.created_at);
CREATE INDEX event_drain_level
  FOR (e:Event) ON (e.drain_level);
CREATE INDEX coding_session_project
  FOR (cs:CodingSession) ON (cs.project_path);
CREATE INDEX coding_session_ended
  FOR (cs:CodingSession) ON (cs.ended_at);
\end{lstlisting}

\section{Autonomous Agent Prompt Template}
\label{app:prompt}

The autonomous agent uses an externalized prompt template stored at \texttt{src/cognitex/prompts/autonomous\_agent.md}. The prompt implements a ``Digital Twin'' persona with state awareness:

\begin{lstlisting}[caption={Digital Twin agent prompt structure (abbreviated)}]
# Digital Twin Autonomous Agent

You are a Digital Twin - an autonomous agent that acts on
behalf of the user when they're not available.

## Your Mission:
1. RESPOND - Draft email replies in the user's voice
2. PREPARE - Compile context packs for upcoming meetings
3. ORGANIZE - Link documents, tasks, and projects
4. SURFACE - Flag truly ambiguous items for human judgment

## The User's Writing Style:
{writing_samples_text}

## LEARNED GUIDELINES & FEEDBACK:
{learned_guidelines}

## User State & Energy Context:
{state_context_text}

## Current State Summary:
- Inbox items needing triage: {inbox_count}
- Emails awaiting response: {emails_needing_response}
- Meetings needing prep: {meetings_needing_prep}
- Connection opportunities: {connection_opportunities}

## STATE-AWARE DECISIONS (Critical):
1. Check User State Context before suggesting tasks
2. In POST-CLINICAL RECOVERY: Only low-energy tasks
3. High fatigue (>70%): Avoid high-friction items
4. Outside peak hours: Use SUGGEST_RESCHEDULE

## Available Actions:
PROCESS_INBOX_ITEM, DRAFT_EMAIL, COMPILE_CONTEXT_PACK,
SCHEDULE_BLOCK, SUGGEST_RESCHEDULE, LINK_DOCUMENT,
LINK_REPOSITORY, LINK_TASK, LINK_PROJECT_TO_GOAL,
CREATE_TASK, FLAG_FOR_REVIEW

## Decision Rules:
1. DRAFT_EMAIL takes priority for actionable emails
2. Prepare context packs for meetings without context
3. Auto-link everything in Connection Opportunities
4. Maximum 5 actions per cycle
5. Maximum 1 FLAG_FOR_REVIEW per cycle
\end{lstlisting}

\section{Agent Action Types}
\label{app:actions}

\begin{table}[h]
\centering
\caption{Autonomous agent action types logged to agent\_actions table.}
\begin{tabular}{ll}
\toprule
\textbf{Action Type} & \textbf{Description} \\
\midrule
\multicolumn{2}{l}{\textit{Digital Twin Actions}} \\
draft\_email & Email reply drafted for user review \\
compile\_context\_pack & Meeting briefing compiled \\
schedule\_block & Focus time block suggested \\
process\_inbox\_item & Firewall inbox item processed \\
\midrule
\multicolumn{2}{l}{\textit{Graph Maintenance Actions}} \\
link\_document & Document linked to project \\
link\_repository & Repository linked to project \\
link\_task & Task linked to project \\
link\_project\_to\_goal & Project linked to goal \\
create\_task & New task created \\
update\_project\_status & Project status changed \\
\midrule
\multicolumn{2}{l}{\textit{State-Aware Actions}} \\
suggest\_reschedule & High-friction task rescheduling suggested \\
flag\_for\_review & Item flagged for human review \\
\midrule
\multicolumn{2}{l}{\textit{System Actions}} \\
autonomous\_cycle & Cycle completion summary \\
drive\_poll & Drive change detection cycle \\
notification\_sent & Discord notification delivered \\
clinical\_recovery & Post-clinical mode activated \\
\bottomrule
\end{tabular}
\end{table}

\section{Operating Modes Reference}
\label{app:modes}

\begin{table}[h]
\centering
\caption{Complete operating mode specifications.}
\begin{tabular}{llllll}
\toprule
\textbf{Mode} & \textbf{Task Types} & \textbf{Friction} & \textbf{Block} & \textbf{Notify} & \textbf{Interrupt} \\
\midrule
DEEP\_FOCUS & all & 5 & 90+ min & batched & urgent, family \\
FRAGMENTED & shallow, admin & 3 & any & all & all \\
OVERLOADED & maintenance, recovery & 1 & any & critical & emergency \\
AVOIDANT & low-friction, quick & 2 & 30+ min & batched & urgent \\
HYPERFOCUS & current only & 5 & 120+ min & none & emergency \\
TRANSITION & shallow, prep & 3 & any & batched & urgent, family \\
\bottomrule
\end{tabular}
\end{table}

\section{Database Schema Extensions}
\label{app:schema_ext}

\begin{lstlisting}[language=SQL,caption={Learning system tables}]
-- State observations for temporal learning
CREATE TABLE state_observations (
    id SERIAL PRIMARY KEY,
    task_id TEXT,
    task_title TEXT,
    outcome TEXT NOT NULL,  -- completed, deferred, abandoned
    mode TEXT,
    fatigue_level FLOAT,
    hour_of_day INTEGER,
    day_of_week INTEGER,
    post_clinical BOOLEAN DEFAULT FALSE,
    energy_cost TEXT,
    observed_at TIMESTAMP DEFAULT NOW()
);

-- Preference rules with lifecycle
CREATE TABLE preference_rules (
    id TEXT PRIMARY KEY,
    rule_name TEXT NOT NULL,
    rule_type TEXT NOT NULL,
    conditions JSONB NOT NULL,
    action JSONB NOT NULL,
    confidence FLOAT DEFAULT 0.5,
    lifecycle TEXT DEFAULT 'candidate',
    applications INTEGER DEFAULT 0,
    successful_applications INTEGER DEFAULT 0,
    success_rate FLOAT,
    created_at TIMESTAMP DEFAULT NOW(),
    validated_at TIMESTAMP,
    deprecated_at TIMESTAMP
);

-- Task proposals for learning
CREATE TABLE task_proposals (
    id TEXT PRIMARY KEY,
    title TEXT NOT NULL,
    project_id TEXT,
    priority TEXT,
    status TEXT DEFAULT 'pending',
    user_feedback TEXT,
    created_at TIMESTAMP DEFAULT NOW(),
    decided_at TIMESTAMP
);
\end{lstlisting}

\end{document}
